\documentclass[epj]{svjour}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{tikz}
\usepackage{xparse}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\usepackage[numbers,sort&compress]{natbib}

\begin{document}

  \title{Bounded Relational Dynamics and the Structural Necessity of Born--Infeld Geometry}

  \author{J.~Beau}

  \institute{Independent Researcher}

  \date{Received: date / Revised version: date}

  \abstract{
    We investigate how spacetime geometry and non-linear field dynamics can arise as
    equilibrium descriptions of relational systems subject to a finite maximal relaxation flux.
    We show that the existence of a universal bound on admissible relational gradients
    excludes purely quadratic effective actions and uniquely enforces a Born--Infeld-type
    structure as the minimal local representation compatible with saturation.
    Starting from a weighted relational Laplacian with irreversible relaxation,
    we derive an effective continuum description in which the metric tensor emerges
    from the principal symbol of the operator, while antisymmetric perturbations
    enter as an effective gauge field strength.
    In homogeneous regimes, the bounded-relaxation constraint dynamically selects
    flat spacetime with pseudo-Riemannian signature $(-+++)$ as a stable equilibrium.
    When homogeneity is broken by a localized stationary obstruction,
    the same mechanism yields the Schwarzschild geometry as the universal effective exterior solution.
    Horizon formation corresponds to saturation of admissible relational flux and to
    a loss of projectability of the continuum description rather than to a physical singularity.
    These results provide an operator-based unification of Lorentzian geometry,
    Born--Infeld electrodynamics, and horizon formation as consequences of bounded relational dynamics.
  }
%
  \PACS{
      {04.20.-q}{Classical general relativity} \and
    {11.10.Lm}{Nonlinear or nonlocal theories and models} \and
    {03.50.Kk}{Field theories in dimensions other than four}
  }

  \maketitle

  \section{Introduction}
    \label{sec:introduction}

    A growing body of work explores the possibility that spacetime geometry is not a
    fundamental structure but an effective description emerging from more primitive
    relational or operator-based systems.
    In several approaches, discrete relational data, weighted graphs, or spectral operators
    are shown to admit continuum limits in which geometric notions such as distance,
    dimension, and curvature arise as derived quantities rather than postulated ones.
    In particular, it has been demonstrated that, under mild regularity and density
    assumptions, the continuum limit of a relational Laplacian converges to a second-order
    elliptic operator whose principal symbol defines an effective metric tensor~\cite{Beau2026a}.

    While such results establish that effective spacetime geometry can emerge from
    non-geometric relational structures, they leave open a central question:
    why are only very specific geometries observed?
    From a purely mathematical perspective, a large class of elliptic operators and
    corresponding metrics are admissible.
    However, not all such operators necessarily correspond to physically meaningful or
    operationally accessible descriptions.
    Recent analyses have emphasized that effective geometric descriptions may fail when
    the underlying relational structure admits non-injective or degenerate projections,
    leading to a loss of observable factorization and to intrinsic limits of spacetime
    representability~\cite{Beau2026b}.

    This work addresses a complementary and more restrictive problem.
    Rather than asking how geometry can emerge, we ask which effective geometries are
    \emph{dynamically admissible} once minimal physical constraints are imposed on the
    underlying relational dynamics.
    Specifically, we consider relational systems whose effective continuum descriptions
    admit a finite maximal propagation or relaxation speed.
    Such a bound is required for causal consistency and excludes purely quadratic actions,
    which allow arbitrarily large gradients and unbounded fluxes.

    We show that imposing bounded flux propagation uniquely constrains the form of any
    local effective functional governing the continuum limit.
    Under mild assumptions of locality, smoothness, and absence of additional microscopic
    scales, the resulting effective description must take a Born--Infeld--type form.
    This structure is not introduced as a phenomenological modification, but arises as the
    minimal representation compatible with flux saturation.

    Building on this result, we demonstrate that flux saturation does more than regularize
    the effective theory.
    It dynamically selects a restricted class of admissible relational Laplacians whose
    continuum limits define stable and physically meaningful metrics.
    In homogeneous regimes, this selection leads uniquely to flat spacetime.
    In the presence of a localized and stationary obstruction, the same mechanism yields
    the Schwarzschild geometry as the universal effective solution, without postulating
    any independent metric dynamics or field equations.

    Within this framework, horizons are interpreted as loci where flux saturation renders
    the effective operator degenerate, signaling a loss of projectability rather than a
    physical singularity.
    This interpretation aligns naturally with structural analyses of non-injective
    projections and clarifies the operational meaning of strong-field regimes.

    The paper is organized as follows.
    In Section~\ref{sec:bounded-relaxation-and-effective-action}, we establish the necessity of a Born--Infeld--type
    structure from bounded flux propagation.
    Section~\ref{sec:rel-graph-laplacian-op-and-relaxation-dynamics} recalls the emergence of effective
    continuum operators from relational Laplacians.
    Section~\ref{sec:continuum-limit-and-effective-operators} discusses metric reconstruction from operator symbols.
    Sections~\ref{sec:emergence-of-minkowski-spacetime} and~\ref{sec:localized-obstruction-and-schwarzschild-geometry}
    derive flat and Schwarzschild geometries as dynamically selected solutions.
    Section~\ref{sec:horizon-as-flux-saturation-and-loss-of-projectability} analyzes horizon formation as operator
    saturation.
    We conclude with a discussion of scope and limitations.

  \section{Bounded relaxation and effective action}
    \label{sec:bounded-relaxation-and-effective-action}

    We consider effective continuum descriptions arising from relational systems whose
    microscopic dynamics admits a finite maximal transport or relaxation flux.
    Such a bound is a minimal structural requirement: the underlying relational substrate
    cannot sustain arbitrarily large effective transport densities without losing
    projectability.

    In a continuum regime, the dynamics of slowly varying configurations can be encoded
    in a local action functional
    \begin{equation}
      S = \int d^4x\,\mathcal{L},
    \end{equation}
    where $\mathcal{L}$ depends on the effective fields and their first derivatives.
    From the perspective of field theory, the central question is therefore the following:
    which local Lagrangian densities are compatible with a finite maximal transport capacity?

    A purely quadratic action, such as the Maxwell form
    \begin{equation}
      \mathcal{L}_{\mathrm{quad}} \propto F_{\mu\nu}F^{\mu\nu},
    \end{equation}
    allows for arbitrarily large field invariants.
    In such a linear theory, the constitutive relation
    $D^{\mu\nu} = \partial \mathcal{L} / \partial F_{\mu\nu}$ remains unbounded.
    Arbitrarily large field invariants therefore correspond to proportionally large
    effective flux densities, without intrinsic saturation.
    This behavior is incompatible with the existence of a finite maximal transport
    capacity at the microscopic relational level.

    Imposing bounded propagation must therefore be understood as imposing bounded
    constitutive response.
    The effective Lagrangian density must satisfy three minimal conditions:
    (i) it reduces to a quadratic form in the weak-field limit,
    (ii) it enforces a strict upper bound on admissible field invariants through its
    constitutive relation,
    and (iii) it introduces no additional microscopic scales beyond the saturation scale
    itself.
    Polynomial extensions of quadratic actions fail to meet these requirements without
    fine tuning or the introduction of auxiliary parameters.

    Under these assumptions, the effective Lagrangian density is uniquely constrained to
    take a Born--Infeld--type form.
    Anticipating the operator derivation developed in the following sections, the resulting
    effective action can be written as
    \begin{equation}
      S_{\mathrm{eff}}
      = \beta^2 \int d^4x
      \left(
        \sqrt{-\det\!\left(g_{\mu\nu} + \frac{1}{\beta} F_{\mu\nu}\right)}
        - \sqrt{-\det(g_{\mu\nu})}
      \right),
      \label{eq:BI-action}
    \end{equation}
    where $g_{\mu\nu}$ is the effective metric and $F_{\mu\nu}$ an antisymmetric field strength.
    The parameter $\beta$ sets the maximal admissible field magnitude and fixes the
    saturation scale of the theory.

    The saturation scale $\beta$ will be referred to as the bounded-flux (Born--Infeld)
    scale.
    It characterizes the maximal admissible relational flux within the validity
    domain of the continuum description and encodes the intrinsic transport capacity
    of the underlying substrate.

    In the weak-field regime $|F_{\mu\nu}| \ll \beta$, the expansion of
    Eq.~\eqref{eq:BI-action} yields
    \begin{equation}
      \mathcal{L}_{\mathrm{eff}}
      = -\frac{1}{4}F_{\mu\nu}F^{\mu\nu}
      + \mathcal{O}\!\left(\beta^{-2}\right),\label{eq:equation-weak-field}
    \end{equation}
    showing that standard Maxwell electrodynamics is recovered universally as the
    low-flux approximation of bounded relaxation.
    Non-linear corrections become relevant only when the microscopic saturation scale
    is probed.

    Crucially, the Born--Infeld structure is not introduced here as a phenomenological
    modification of electrodynamics.
    It arises as the minimal local representation compatible with bounded constitutive
    response.
    This statement can be formulated as a structural no-go result:
    any local effective action admitting a finite maximal flux cannot remain purely
    quadratic in field gradients.
    Once locality, Lorentz compatibility, analyticity around the weak-field limit,
    and the absence of additional microscopic scales are imposed,
    the Born--Infeld class provides the unique analytic completion ensuring saturation
    of the constitutive relation.

    \begin{figure}[t]
      \centering
      \begin{tikzpicture}
        \begin{axis}[
          width=0.85\columnwidth,
          height=0.55\columnwidth,
          xlabel={Field gradient or intensity},
          ylabel={Propagated flux},
          xmin=0, xmax=4,
          ymin=0, ymax=1.4,
          samples=200,
          axis lines=left,
          legend style={at={(0.97,0.35)},anchor=east},
          ticklabel style={font=\small},
          label style={font=\small}
        ]

          \addplot[thick, dashed]
          {0.35*x};
          \addlegendentry{Linear flux (Maxwell)}

          \addplot[thick]
          {1 - exp(-0.9*x)};
          \addlegendentry{Bounded flux (Born--Infeld)}

        \end{axis}
      \end{tikzpicture}
      \caption{
        Comparison between linear and bounded flux propagation.
        Quadratic actions yield unbounded constitutive response,
        whereas bounded relaxation enforces intrinsic saturation,
        naturally selecting a Born--Infeld--type structure.
      }
      \label{fig:bounded-vs-linear-flux}
    \end{figure}

    In the present framework, the parameter $\beta$ is not free.
    It is determined by the maximal transport capacity and connectivity density of the
    underlying relational substrate.
    Consequently, this scale is universal across all effective sectors emerging from the
    same operator description.
    Both geometric and gauge degrees of freedom therefore share a common saturation
    threshold set by the substrate itself.

    The remainder of the paper establishes how the effective action
    \eqref{eq:BI-action} emerges dynamically from a bounded relational Laplacian and how
    this structure selects physically admissible spacetime geometries.

  \section{Structural Necessity of Non-Injective Projection}
    \label{sec:noninjective-necessity}

    The bounded relaxation principle introduced in Section~2 was formulated as a structural
    property of the effective description.
    We now show that this boundedness is not an independent assumption, but a necessary
    consequence of requiring emergent gauge structure within a relational projection framework.

    \subsection{Admissible projections and holonomy}
      \label{subsec:admissible-projections-and-holonomy}

      Let $\Pi : \Omega \to \mathcal{O}$ denote the projection from the relational configuration
      space $\Omega$ to effective observables $\mathcal{O}$.
      We assume that admissible projections satisfy locality and compatibility with the
      relaxation ordering introduced in Section~2.

      \begin{proposition}
        If $\Pi$ is injective and compatible with relational locality,
        then the induced connection on the projected fiber bundle has trivial holonomy.
      \end{proposition}

      \begin{proof}[Sketch]
        Injectivity implies that each observable configuration corresponds to a unique
        relational configuration.
        Parallel transport of effective states along any closed loop in $\mathcal{O}$
        lifts uniquely to a closed loop in $\Omega$.
        Since no two distinct relational configurations are identified,
        there exists no non-trivial re-identification symmetry.
        Therefore the holonomy group reduces to the identity element.
      \end{proof}

      \begin{corollary}
        An injective admissible projection cannot generate an emergent gauge structure.
      \end{corollary}

      Gauge symmetry requires the existence of non-trivial fiber re-identifications.
      Trivial holonomy excludes such internal degrees of freedom.
      Therefore, if electromagnetism is to arise as an effective $U(1)$ fiber symmetry,
      the projection $\Pi$ must be non-injective.

      \paragraph{Remark on structural necessity.}
        The necessity of non-injective projection invoked here is not a modeling assumption.
        A general classification of admissible projections shows that any injective projection
        compatible with relational locality has trivial holonomy and therefore cannot support
        emergent gauge or curvature structure.
        Non-injectivity is thus a structural requirement for the emergence of internal fiber
        symmetries.
        A complete proof of this classification result is given in~\cite{Beau2026whitepaper}.

  \subsection{Finite transport capacity from non-injectivity}
    \label{subsec:finite-transport-capacity-from-non-injectivity}

    Non-injectivity implies that multiple relational configurations are mapped
    to the same effective observable.
    Let $\mathcal{E}_y = \Pi^{-1}(y)$ denote the equivalence class associated with
    an observable $y \in \mathcal{O}$.

    Because $\mathcal{E}_y$ is non-trivial,
    the projection cannot resolve arbitrarily fine variations of relational flux.
    Transport of relaxation across the projection therefore admits a maximal
    admissible rate.
    Beyond this rate, distinct relational configurations collapse into the same
    effective description, and injectivity fails locally.

    \begin{proposition}
      Non-injective projection implies the existence of a finite maximal
      effective transport capacity.
    \end{proposition}

    This bound is not imposed externally.
    It follows from the finite projectability of relational information
    through $\Pi$.
    The effective field strength cannot grow without limit,
    since sufficiently large gradients correspond to unresolved relational
    distinctions.

  \subsection{Selection of the Born--Infeld completion}
    \label{subsec:selection-of-the-born--infeld-completion}

    Consider a low-gradient regime in which the effective Lagrangian admits
    a quadratic expansion
    \begin{equation}
      \mathcal{L}_{\mathrm{eff}} = -\frac{1}{4}F_{\mu\nu}F^{\mu\nu}
      + \mathcal{O}(F^4).
    \end{equation}

    If transport capacity is finite, the Lagrangian density must remain bounded
    for large invariants $F_{\mu\nu}F^{\mu\nu}$.
    Analyticity and Lorentz invariance then restrict the admissible nonlinear
    completions.

    Under these conditions, the minimal deformation that:
    \begin{enumerate}
      \item preserves gauge invariance,
      \item reduces to Maxwell theory at low field strength,
      \item bounds the field invariants,
    \end{enumerate}
    is of the Born--Infeld type,
    \begin{equation}
      \mathcal{L}_{\mathrm{BI}} =
      \beta^2 \left(
                1 - \sqrt{1 + \frac{1}{2\beta^2}F_{\mu\nu}F^{\mu\nu}
          - \frac{1}{16\beta^4}(F_{\mu\nu}\tilde F^{\mu\nu})^2}
      \right),
    \end{equation}
    where $\beta$ sets the saturation scale.

    \begin{theorem}
      If emergent electromagnetism arises from a relational projection
      that is local, Lorentz-compatible, and non-injective,
      then the effective nonlinear completion of Maxwell theory must
      belong to the Born--Infeld class.
    \end{theorem}

    \begin{proof}[Sketch]
      Non-injectivity implies finite transport capacity.
      Finite capacity requires bounded field invariants.
      Gauge invariance and analyticity restrict admissible
      Lorentz scalars to functions of $F^2$ and $(F\tilde F)^2$.
      The Born--Infeld structure is the unique analytic completion
      that preserves Maxwell behavior at low field strength
      while ensuring bounded invariants.
    \end{proof}

  \subsection{Interpretational consequence}
    \label{subsec:interpretational-consequence}

    Born--Infeld dynamics is therefore not an arbitrary ultraviolet
    modification.
    It is structurally selected once three conditions are imposed:
    \begin{enumerate}
      \item emergent gauge symmetry,
      \item relational locality,
      \item non-injective projection.
    \end{enumerate}

    The bounded relaxation principle introduced in Section~2
    is thus a necessary consequence of the geometric structure of
    the projection itself.
    In the following sections, this structural saturation will be shown
    to control both spectral stability and horizon formation.

  \section{Relational graph, Laplacian operator, and relaxation dynamics}
  \label{sec:rel-graph-laplacian-op-and-relaxation-dynamics}

  We consider an underlying relational system represented by a weighted graph
  $G = (V,E,w)$, where $V$ denotes a set of nodes, $E$ a set of edges, and
  $w_{ij} \ge 0$ encodes the strength of the relation between nodes $i$ and $j$.
  No embedding in a background spacetime is assumed.
  The graph is taken as a purely relational object, encoding adjacency and coupling
  structure between abstract degrees of freedom.

  A scalar field $\phi_i$ is defined on the nodes of the graph.
  The fundamental operator governing relational variations is the discrete Laplacian,
  defined by
  \begin{equation}
    (L \phi)_i = \sum_{j} w_{ij} \left( \phi_i - \phi_j \right) .
  \end{equation}
  This operator measures the local mismatch of $\phi$ with respect to its relational
  neighborhood and plays the role of a generalized stiffness or connectivity operator.
  Throughout this work, we assume that the graph is locally finite and that the weights
  are symmetric, $w_{ij} = w_{ji}$.

  The system is assumed to admit an irreversible relaxation dynamics driven by the
  Laplacian.
  At the discrete level, this dynamics can be represented schematically as
  \begin{equation}
    \frac{d \phi_i}{d \tau} = - (L \phi)_i ,
  \end{equation}
  where $\tau$ is a monotonically increasing parameter labeling the progression of the relaxation process.
  No interpretation of $\tau$ as a fundamental physical time is imposed.
  Instead, it serves as an ordering parameter associated with the irreversible flow
  toward admissible stationary configurations.

  This relaxation dynamics defines a preferred direction of evolution.
  Configurations evolve toward states that minimize relational gradients, subject to
  the constraints discussed in Section~\ref{sec:bounded-relaxation-and-effective-action}.
  The existence of such an ordering parameter is sufficient to distinguish one direction
  of evolution from its reverse, independently of any geometric notion of time.
  Temporal ordering is thus introduced operationally, as an intrinsic feature of the
  relaxation process itself.

  Stationary configurations satisfy
  \begin{equation}
    L \phi = 0 ,
  \end{equation}
  and correspond to relational equilibria.
  More generally, slowly varying configurations describe regimes in which the system
  admits an effective coarse-grained description.
  In these regimes, large-scale observables depend primarily on the spectral properties
  of the Laplacian rather than on the detailed microscopic structure of the graph.

  We emphasize that the graph structure introduced here does not imply any fundamental
  discreteness of physical space or time.
  It is employed as a relational scaffold allowing the definition of operators and
  relaxation processes.
  Different microscopic realizations leading to the same large-scale spectral properties
  are considered physically equivalent within the scope of the effective description.

  In the next section, we recall how, under appropriate density and regularity
  assumptions, the spectral structure of such relational Laplacians admits a continuum
  limit.
  In this limit, the discrete operator converges to a second-order differential operator,
  providing the bridge toward an effective geometric description.

  \section{Continuum limit and Born--Infeld action}
  \label{sec:continuum-limit-and-effective-operators}

  We briefly recall how an effective continuum description and its associated action
  emerge from the spectral structure of a bounded relational Laplacian.
  Detailed proofs of the continuum limit and operator convergence can be found
  in~\cite{Beau2026a}.
  Here we summarize only the elements required to establish the effective
  field-theoretic description in a form sufficient for the subsequent analysis.

  We consider a sequence of weighted graphs
  $G_N=(V_N,E_N,w^{(N)})$ with increasing cardinality $|V_N|\to\infty$.
  The graphs are assumed to satisfy three structural conditions:
  (i) uniformly bounded degree growth relative to the intrinsic scaling,
  (ii) local quasi-isotropy in the sense that weighted second moments of adjacency
  converge to an isotropic tensor in the low-energy sector,
  and (iii) sufficient decay of $w_{ij}^{(N)}$ with intrinsic relational distance to
  ensure locality of the limit operator.
  No background embedding is assumed.
  All notions of proximity are defined intrinsically from the graph structure.

  Let $\phi_i$ denote a scalar degree of freedom on the nodes of $G_N$.
  The discrete Laplacian acts as
  \begin{equation}
    (L_N \phi)_i = \sum_j w_{ij}^{(N)}\left(\phi_i-\phi_j\right).
  \end{equation}

  Under the above conditions, the family $L_N$ converges in the strong resolvent
  sense, within a fixed low-energy spectral window, to a second-order differential
  operator on a smooth manifold $\mathcal{M}$.
  More precisely, for modes whose eigenvalues remain bounded as $N\to\infty$,
  the action of $L_N$ on slowly varying configurations converges to
  \begin{equation}
    L_N \;\longrightarrow\;
    \mathcal{L}
    = \nabla_\mu\!\left(A^{\mu\nu}(x)\nabla_\nu\right),
  \end{equation}
  where $A^{\mu\nu}(x)$ is a symmetric tensor encoding the local connectivity
  and stiffness of the underlying relational system.

  The convergence holds in the sense that resolvents
  $(L_N+\lambda I)^{-1}$ converge strongly to
  $(\mathcal{L}+\lambda I)^{-1}$ for $\lambda$ in the resolvent set
  associated with the low-energy sector.
  This guarantees stability of the principal symbol under bounded perturbations
  of the microscopic graph structure and ensures that the effective operator is
  insensitive to microscopic rearrangements that do not alter the spectral window.

  The tensor $A^{\mu\nu}(x)$ arises from the second weighted moments of adjacency.
  Local quasi-isotropy implies that anisotropic corrections remain controlled
  and contribute only higher-order perturbations to the principal symbol.
  Consequently, the leading-order propagation properties of continuum modes are
  robust under bounded microscopic anisotropies.

  In admissible continuum regimes, the operator $\mathcal{L}$ is elliptic on
  spatial slices and admits a well-defined principal symbol,
  \begin{equation}
    \sigma_2(\mathcal{L})(x,k)=A^{\mu\nu}(x)k_\mu k_\nu,
  \end{equation}
  which fully characterizes the leading-order propagation of modes.
  As shown in~\cite{Beau2026a}, this symbol provides a natural,
  coordinate-independent definition of an effective metric tensor,
  \begin{equation}
    g^{\mu\nu}(x)\propto A^{\mu\nu}(x),
  \end{equation}
  where the proportionality factor reflects a choice of units.

  Crucially, the effective metric $g_{\mu\nu}$ is not introduced as an independent
  dynamical variable.
  It is a derived object encoding how relational variations propagate in the
  continuum limit.
  Distinct microscopic graphs leading to the same principal symbol are therefore
  indistinguishable at the level of effective geometry.
  The continuum description is thus defined only up to spectral equivalence
  within the selected energy window.

  When antisymmetric perturbations of the relational connectivity are included,
  the continuum operator acquires an additional two-form contribution that
  naturally enters as a field strength $F_{\mu\nu}$.
  As discussed in
  Section~\ref{sec:bounded-relaxation-and-effective-action},
  bounded relaxation constrains the admissible gradients of these perturbations.
  At the level of the effective continuum description, this constraint is most
  consistently implemented at the level of the action rather than the equations
  of motion.

  The resulting effective action takes a Born--Infeld form,
  \begin{equation}
    S_{\mathrm{eff}}
    = \beta^2 \int d^4x
    \left(
      \sqrt{-\det\!\left(g_{\mu\nu}+\frac{1}{\beta}F_{\mu\nu}\right)}
      -\sqrt{-\det(g_{\mu\nu})}
    \right),
    \label{eq:BI-action-continuum}
  \end{equation}
  where $\beta$ sets the universal maximal admissible relational flux.
  In the weak-field regime, this action reduces to the Maxwell form,
  while at large field strengths it enforces saturation of the constitutive
  response.

  The validity of this effective action is restricted to regimes in which the
  spectrum of $L_N$ admits a stable low-energy sector and the bounded-flux
  condition remains satisfied.
  Outside this spectral window, strong-resolvent convergence no longer holds,
  the principal symbol ceases to provide a faithful description,
  and geometric or field-theoretic notions lose operational meaning.

  In the following section, we show that the bounded-relaxation constraint further
  restricts the admissible forms of $A^{\mu\nu}(x)$.
  In homogeneous regimes, these restrictions uniquely select a flat effective
  geometry with a specific signature.

  \section{Emergence of Minkowski Spacetime: Spectral Stability and Signature Selection}
  \label{sec:emergence-of-minkowski-spacetime}

  We now demonstrate that the class of homogeneous and isotropic relaxation regimes
  uniquely selects a flat pseudo-Riemannian metric.
  In this framework, homogeneity is defined spectrally: the low-energy sector of the
  relational Laplacian $\mathcal{L}$ is invariant under the translation group,
  implying that the effective tensor $A^{\mu\nu}(x)$ reduces to a constant matrix
  $A^{\mu\nu}_0$.

  The selection of the metric signature follows from a structural compatibility
  requirement between irreversible relaxation, bounded constitutive response, and
  spectral stability of the effective operator.
  Let $\tau$ denote the monotonic ordering parameter associated with relational
  relaxation.
  The effective continuum dynamics must admit a well-posed initial value problem
  relative to $\tau$ in the regime where the projection remains local and injective.

  The bounded-flux condition established in
  Section~\ref{sec:bounded-relaxation-and-effective-action}
  excludes purely diffusive effective descriptions.
  A parabolic operator would imply instantaneous spreading of perturbations across
  the relational network, as is characteristic of elliptic spatial operators coupled
  to first-order diffusion in $\tau$.
  Such behavior is incompatible with a finite maximal transport capacity, since it
  destroys any notion of a finite characteristic cone for effective propagation.

  In the homogeneous regime, the principal symbol of the effective operator is
  determined by $A^{\mu\nu}_0$.
  Denoting $k_0$ the frequency conjugate to $\tau$ and $\mathbf{k}$ the spatial
  momenta, spectral stability requires that the dispersion relation define a
  hyperbolic operator in the sense of Hörmander~\cite{HormanderPDE}.
  Hyperbolicity ensures the existence of a finite characteristic cone and a
  well-posed Cauchy problem for effective continuum modes.

  A purely Riemannian signature $(++++)$ would render the principal symbol
  positive definite.
  The corresponding operator is elliptic, not hyperbolic.
  No finite propagation cone exists in this case, and perturbations spread
  instantaneously in the effective description, violating the bounded-transport
  requirement.

  Conversely, a signature with more than one time-like direction leads to
  ultra-hyperbolic equations.
  Such operators generically fail to admit a globally well-posed initial value
  problem and are unstable under small perturbations of initial data~\cite{ChoquetBruhat2009}.
  They therefore cannot provide a spectrally stable fixed point of the relaxation
  dynamics.

  The only configuration compatible with
  (i) hyperbolicity of the principal symbol,
  (ii) existence of a finite characteristic cone,
  (iii) monotonic ordering along $\tau$,
  and (iv) bounded constitutive response,
  is the case in which the ordering direction carries a sign opposite to the
  spatial directions.
  After normalization of the maximal relaxation speed to $c=1$, the effective
  metric is dynamically fixed to
  \begin{equation}
    g_{\mu\nu} = \eta_{\mu\nu} = \mathrm{diag}(-1, +1, +1, +1).
  \end{equation}

  Lorentz invariance thus emerges as a symmetry of the homogeneous spectral fixed
  point.
  The Minkowski metric is not postulated as a background structure.
  It is the unique effective representation of a maximally symmetric relational
  equilibrium compatible with bounded relaxation.

  The emergence of the $(-+++)$ signature also clarifies the role of Wick rotation.
  The Euclidean sector corresponds to relational connectivity on a fixed relaxation
  slice, where the operator is effectively elliptic.
  The Lorentzian signature encodes the dynamical unfolding of relational structure
  along the monotonic ordering direction.
  The null interval $ds^2=0$ corresponds to saturation of the effective characteristic
  cone, where relational transport reaches the maximal rate permitted by the
  microscopic connectivity of $\chi$.
  This mechanism parallels the causal saturation encountered in non-linear
  Born--Infeld-type field theories~\cite{Gibbons1998}.

  In the absence of localized obstructions, all curvature invariants
  $R_{\mu\nu\rho\sigma}$ vanish identically.
  The homogeneous relational fixed point therefore corresponds to flat spacetime.
  In the next section, we show how localized defects break this maximal symmetry
  and generate the Schwarzschild geometry through the same bounded-saturation
  mechanism.

  \section{Localized obstruction and Schwarzschild geometry}
  \label{sec:localized-obstruction-and-schwarzschild-geometry}

  We now consider relaxation regimes in which homogeneity is broken by the presence of a
  localized and stationary obstruction.
  Operationally, such an obstruction corresponds to a region where the relational
  connectivity or stiffness of the underlying system is enhanced, thereby constraining
  the local relaxation dynamics.
  No assumption is made regarding the microscopic origin of this obstruction.
  Only its macroscopic effect on the effective operator is considered.

  Outside the obstructed region, the system remains stationary and isotropic.
  The effective continuum operator introduced in
  Section~\ref{sec:continuum-limit-and-effective-operators}
  therefore takes the form
  \begin{equation}
    \mathcal{L}
    = \nabla_\mu \!\left( A^{\mu\nu}(r)\,\nabla_\nu \right) ,
  \end{equation}
  where $r$ denotes the radial distance from the center of the obstruction.
  Spherical symmetry implies that $A^{\mu\nu}(r)$ depends only on $r$ and decomposes into
  temporal and spatial components,
  \begin{equation}
    A^{\mu\nu}(r) =
    \mathrm{diag}\!\left(-A_t(r),\,A_r(r),\,A_\perp(r),\,A_\perp(r)\right) .
  \end{equation}
  The sign structure is inherited from the hyperbolic principal symbol selected in
  Section~\ref{sec:emergence-of-minkowski-spacetime}.

  In the exterior region, where no additional sources are present, the relaxation dynamics
  is governed by the homogeneous equation
  \begin{equation}
    \nabla_\mu \!\left( A^{\mu\nu}(r)\,\nabla_\nu \Phi \right) = 0 ,
  \end{equation}
  where $\Phi$ represents a slowly varying scalar probe of the effective structure.
  Under stationarity and spherical symmetry, this equation reduces to
  \begin{equation}
    \frac{1}{r^{2}}
    \frac{d}{dr}
    \left(
      r^{2} A_r(r) \frac{d\Phi}{dr}
    \right)
    = 0 .
  \end{equation}
  This relation expresses conservation of effective flux through spheres of radius $r$.
  Its general solution is
  \begin{equation}
    \Phi(r) = \Phi_0 - \frac{C}{r} ,
  \end{equation}
  where $C$ is an integration constant characterizing the strength of the obstruction.

  The emergence of the $1/r$ profile is therefore not imposed by symmetry alone.
  It arises within the class of effective operators already constrained by bounded
  constitutive response and Lorentz-compatible hyperbolicity.
  In particular, it holds provided that
  (i) the continuum projection remains local and injective in the exterior region,
  (ii) effective isotropy is preserved,
  and (iii) no additional microscopic scales are introduced beyond the universal
  saturation scale defined in Section~\ref{sec:bounded-relaxation-and-effective-action}.
  Under these conditions, flux conservation uniquely fixes the radial decay.

  Using the identification between the principal symbol of $\mathcal{L}$ and the effective
  metric established in
  Section~\ref{sec:continuum-limit-and-effective-operators},
  the radial dependence of $A^{\mu\nu}(r)$ translates directly into a
  position-dependent metric tensor.
  Up to a choice of coordinates consistent with stationarity and spherical symmetry,
  the effective line element can be written as
  \begin{equation}
    ds^{2}
    = - f(r)\, dt^{2}
    + f(r)^{-1} dr^{2}
    + r^{2} d\Omega^{2} ,
  \end{equation}
  where the lapse function $f(r)$ is determined by the radial profile of the operator
  coefficients.
  Matching the weak-obstruction limit with the homogeneous solution of
  Section~\ref{sec:emergence-of-minkowski-spacetime}
  fixes
  \begin{equation}
    f(r) = 1 - \frac{r_s}{r} ,
  \end{equation}
  where $r_s$ is a constant proportional to $C$ and measures the integrated flux
  associated with the obstruction.

  The resulting geometry coincides with the Schwarzschild metric.
  Here, however, it is not obtained from an independent set of metric field equations.
  It arises as the universal stationary fixed point of a bounded relaxation operator
  in the presence of a localized and isotropic perturbation.
  The parameter $r_s$ characterizes the strength of the obstruction in relational units
  and is identified operationally through its influence on admissible propagation cones
  and relaxation rates.

  This derivation highlights that the Schwarzschild geometry is a universal effective
  response within a bounded-relaxation framework.
  Its form is fixed by symmetry, stationarity, hyperbolicity of the principal symbol,
  and conservation of saturated flux.
  No independent curvature dynamics or energy--momentum tensor needs to be postulated
  at the effective level.
  The geometry encodes how admissible continuum modes propagate when relaxation is
  constrained by a localized enhancement of relational stiffness.

  In the next section, we analyze the behavior of the effective operator near the radius
  $r = r_s$.
  We show that this surface corresponds to saturation of the bounded-transport condition,
  where the principal symbol degenerates and the continuum chart loses projectability,
  providing an operator-theoretic interpretation of horizons.

  \section{Horizon as flux saturation and loss of projectability}
  \label{sec:horizon-as-flux-saturation-and-loss-of-projectability}

  We now examine the behavior of the effective continuum description in the vicinity of
  the radius $r=r_s$ identified in
  Section~\ref{sec:localized-obstruction-and-schwarzschild-geometry}.
  At this radius, the lapse function $f(r)$ vanishes and the effective geometry develops
  a horizon in the standard geometric representation.
  We show that, in the present framework, this phenomenon admits a precise
  operator-theoretic interpretation in terms of saturation of the principal symbol.

  As established in
  Section~\ref{sec:bounded-relaxation-and-effective-action},
  the effective action and its associated continuum operator remain valid only as long
  as the bounded constitutive response is satisfied.
  In the presence of a localized obstruction, the radial dependence of the operator
  coefficients $A^{\mu\nu}(r)$ reflects the increasing constraint imposed on relational
  relaxation.
  As $r$ decreases, admissible gradients approach the maximal value set by the universal
  saturation scale $\beta$.
  At $r=r_s$, this bound is reached in the effective description.

  At this radius, the principal symbol of the operator
  \[
    \sigma_2(\mathcal{L})(x,k)
    = A^{\mu\nu}(r)\,k_\mu k_\nu
  \]
  becomes degenerate.
  Equivalently, the determinant of $A^{\mu\nu}(r)$ vanishes at $r=r_s$.
  The operator therefore undergoes a local change of type.
  Hyperbolicity, which guaranteed a well-posed Cauchy problem in the exterior region,
  is lost at this surface.

  This loss of hyperbolicity implies that the effective continuum description can no
  longer sustain propagating modes with a finite characteristic cone.
  The Cauchy problem for effective fields ceases to be well posed at $r=r_s$.
  Importantly, this degeneracy does not signal a singularity of the underlying
  relational substrate.
  It indicates only the breakdown of the continuum chart obtained through projection.

  From the perspective of the effective action, the horizon corresponds to the locus
  where Born--Infeld saturation becomes operative in the constitutive relation.
  The non-linear structure of the action enforces a maximal admissible flux density,
  preventing divergences in field invariants and energy densities.
  As in standard Born--Infeld electrodynamics, saturation regularizes quantities that
  would otherwise diverge in a purely quadratic theory.

  From the viewpoint of the relational dynamics, the horizon separates two distinct
  regimes of projectability.
  For $r>r_s$, the relaxation dynamics admits a locally injective projection into a
  continuum description with well-defined geometric observables.
  At $r=r_s$, distinct microscopic configurations begin to map to identical effective
  operator data.
  The projection from relational states to spacetime observables becomes
  intrinsically non-injective.

  A quantitative measure of this degeneracy can be expressed through a projection
  entropy defined by the multiplicity of underlying configurations compatible with a
  given effective observable:
  \begin{equation}
    S_{\Pi}
    = -\sum_{o \in O}
    \mu(\Pi^{-1}(o)) \log \mu(\Pi^{-1}(o)).
    \label{eq:projection-entropy}
  \end{equation}
  Horizon formation corresponds to the regime in which the growth of
  $\Pi^{-1}(o)$ limits the local resolvability of continuum observables.

  This interpretation is consistent with structural analyses of non-injective
  effective descriptions, in which loss of observable factorization reflects a
  limitation of the effective representation rather than a breakdown of the
  microscopic dynamics~\cite{Beau2026b}.
  In the present framework, the horizon is precisely the surface where bounded
  transport and hyperbolicity cease to be simultaneously compatible.

  No extension of the effective spacetime manifold beyond $r_s$ is required for
  internal consistency.
  The continuum description is explicitly restricted to regimes in which the
  principal symbol remains non-degenerate.
  The horizon marks the boundary of this domain of applicability.

  This operator-based interpretation reframes horizons as kinematic consequences
  of bounded relaxation rather than as geometric singularities.
  It explains the universality of horizon formation within the class of bounded
  relational operators and clarifies why attempts to probe beyond the horizon
  using effective spacetime observables are intrinsically limited by the
  saturation of the constitutive response.

  \section{Discussion and conclusion}
  \label{sec:discussion-and-conclusion}

  In this work, we have investigated how effective spacetime geometries and
  non-linear field dynamics can emerge from a relational system subject to
  bounded transport.
  Starting from a weighted relational graph endowed with an irreversible
  relaxation process, we have shown that minimal structural constraints
  severely restrict the class of admissible continuum descriptions.

  A first central result is that bounded constitutive response excludes purely
  quadratic effective actions.
  Once locality, analyticity around the weak-field regime, Lorentz compatibility
  of the principal symbol, and the absence of additional microscopic scales are
  imposed, the Born--Infeld class provides the unique analytic completion
  compatible with saturation.
  In this sense, Born--Infeld electrodynamics does not appear as a
  phenomenological deformation of Maxwell theory, but as the minimal local
  structure enforcing a finite maximal flux.
  Standard Maxwell dynamics is recovered universally as the weak-field limit
  of this saturated regime.

  Building on established results concerning the continuum limit of relational
  Laplacians~\cite{Beau2026a}, we have shown that bounded relaxation constrains
  not only the action but also the admissible class of effective differential
  operators.
  In the low-energy spectral window where strong-resolvent convergence holds,
  the principal symbol of the limiting operator defines an effective metric
  tensor.
  In homogeneous and quasi-isotropic relaxation regimes, spectral stability
  and hyperbolicity uniquely select flat spacetime with pseudo-Riemannian
  signature $(- + + +)$.
  Minkowski geometry thus arises as a dynamically admissible fixed point of
  bounded relational relaxation rather than as a fundamental background.

  When homogeneity is broken by a localized and stationary obstruction,
  conservation of saturated flux within the hyperbolic operator class
  generically produces a $1/r$ relaxation profile.
  Through the operator--metric correspondence, this profile induces the
  Schwarzschild geometry as the universal effective description of the exterior
  region.
  This result follows from symmetry, stationarity, hyperbolicity of the
  principal symbol, and bounded transport, without postulating independent
  metric field equations.

  Horizons admit a precise operator-theoretic interpretation.
  They correspond to surfaces where the principal symbol degenerates and
  hyperbolicity is locally lost.
  The associated failure of the Cauchy problem signals a breakdown of the
  effective continuum chart, not a singularity of the underlying relational
  substrate.
  This interpretation is consistent with structural analyses of non-injective
  effective descriptions~\cite{Beau2026b} and clarifies the operational meaning
  of strong-field regimes.

  Saturation regimes also admit a complementary interpretation in terms of
  projection-limited descriptions.
  As the constitutive response approaches its maximal value, distinct
  microscopic relational configurations become indistinguishable at the level
  of effective operators.
  Rather than disappearing, unresolved structure is re-encoded into effective
  geometric or thermodynamic parameters that compensate for the reduced
  local resolvability of the continuum representation.

  The scope of the present work is deliberately limited.
  We do not propose a complete microscopic theory of gravitation, nor do we
  specify the fundamental origin of the relational dynamics.
  Our results apply only within regimes where a low-energy spectral sector
  exists, strong-resolvent convergence holds, and bounded relaxation is
  satisfied.
  Outside this domain, geometric and field-theoretic notions need not remain
  valid, and no extension of the effective spacetime manifold is implied.

  Within these limits, the framework provides a unified structural explanation
  for the emergence of flat spacetime, Schwarzschild geometry, and horizon
  formation as dynamically selected effective structures.
  It suggests that key features traditionally attributed to gravitational
  dynamics may reflect universal properties of bounded relational transport
  rather than independent curvature degrees of freedom.
  Further work will be required to analyze non-stationary regimes,
  departures from spherical symmetry, and possible phenomenological
  consequences of this operator-based perspective.

  \section*{Declaration of generative AI and AI-assisted technologies in the manuscript preparation process}
  During the preparation of this work, the author used generative AI tools
  as an auxiliary support for language refinement, structural clarification,
  and consistency checking.
  After using these tools, the author reviewed and edited the content as needed
  and takes full responsibility for the scientific content, interpretations,
  and conclusions of the published article.

  \bibliographystyle{unsrt}
  \bibliography{references}

\end{document}
